Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-07-27 doc/thesis/acknowledgements.tex
Defence version of thesis with changes

First and foremost I am deeply indebted to my supervisor, Dr. Art Poon, 
for mentoring me through the majority of my academic career.

I also thank my labmates, especially Jeff, Richard, and Thuy. I'll miss our
Thursday lunches.

I am grateful for the mentorship of my committee members Dr. Alexandre
Bouchard-C\^ot\'e, Dr. Richard Harrigan, and Dr. Sarah Otto. Thanks also to my
rotation supervisors Dr. Sara Mostafavi and Dr. Ryan Morin, who accepted me
into their labs and always had time for my many questions.

I owe thanks to my fellow graduate students of the bioinformatics training
program, especially to Ogan.

>>>>

First and foremost I am deeply indebted to my supervisor, Dr. Art Poon, for
mentoring me through the majority of my academic career, and somehow always
managing to make time for me. My time in his lab has been challenging and
rewarding, and I feel well prepared for opportunities to come. I also thank my
labmates, especially Jeff, Richard, and Thuy. I'll miss our Thursday lunches.

I am grateful for the guidance of my committee members Dr. Alexandre
Bouchard-C\^ot\'e, Dr. Richard Harrigan, and Dr. Sarah Otto. Thanks also to my
rotation supervisors Dr. Sara Mostafavi and Dr. Ryan Morin, who accepted me
into their labs and always had time for my many questions.

I owe thanks to my fellow graduate students of the bioinformatics training
program, especially Ogan, who helped keep me on an even keel when times got
tough.

### 2016-07-27 doc/thesis/appendix.Rnw
Defence version of thesis with changes

\begin{figure}[ht]
    \centering
    \includegraphics{abc-point-estimate-m3.pdf}
    \caption[
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        \gls{m} = 3. 
    ]{
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        \gls{m} = 3. Dashed lines indicate true values. (A) Estimates of
        \gls{alpha} and \gls{I} which were varied in these simulations against
        known values. (B) Estimates of \gls{m} and \gls{N} which were held
        fixed in these simulations at the values \gls{m} = 3 and \gls{N} =
        5000.
    }        
    \label{fig:abcptm3}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{abc-point-estimate-m4.pdf}
    \caption[
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        \gls{m} = 3.
    ]{
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        \gls{m} = 4. Dashed lines indicate true values. (A) Estimates of
        \gls{alpha} and \gls{I} which were varied in these simulations against
        known values. (B) Estimates of \gls{m} and \gls{N} which were held
        fixed in these simulations at the values \gls{m} = 4 and \gls{N} =
        5000.
    }        
    \label{fig:abcptm4}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{mixed-posterior}
    \caption[
        Approximate marginal posterior distributions of \gls{BA} model
        parameters obtained with \software{netabc} for a network with
        heterogeneous node behaviour.
    ]{
        Approximate marginal posterior distributions of \gls{BA} model
        parameters obtained with \software{netabc} for a network with
        heterogeneous node behaviour. Half of the nodes were attached with
        $\gls{alpha} = 0.5$, and the other half with $\gls{alpha} = 1.5$
        (vertical dashed lines, top left). Other parameter values were $m = 2$,
        $\gls{I} = 1000$, and $\gls{N} = 5000$ (vertical dashed lines, other
        than top left). Shaded areas indicate 95\% highest posterior density
        intervals.
    }
    \label{fig:mixed}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{peerdriven-posterior}
    \caption[
        Approximate marginal posterior distributions of \gls{BA}
        model parameters obtained with \software{netabc} for a network with
        peer-driven sampling.
    ]{
        Approximate marginal posterior distributions of \gls{BA}
        model parameters obtained with \software{netabc} for a network with
        peer-driven sampling. An epidemic was simulated in the usual fashion,
        but rather than being sampled at random, infected nodes were sampled
        with a probability two times higher if they had any sampled neighbours
        in the contact network. Vertical dashed lines indicate true parameter
        values, and shaded areas indicate 95\% highest posterior density
        intervals.
    }
    \label{fig:peerdriven}
\end{figure}

\begin{figure}[ht]
    \includegraphics{realdata-hpd-bc-m2}
    \caption[
        Posterior mean point estimates and 95\% \gls{HPD} intervals for
        parameters of the \gls{BA} network model, fitted to eleven \gls{HIV}
        datasets with \software{netabc} using the prior \gls{m} $\sim$
        DiscreteUniform(2, 5).
    ]{
        Posterior mean point estimates and 95\% \gls{HPD} intervals for
        parameters of the \gls{BA} network model, fitted to five published
        \gls{HIV} datasets with \software{netabc} using the prior $\gls{m}
        \sim \DiscreteUniform(2, 5)$. $x$-axes indicate regions of nonzero
        prior density. 
    }
    \label{fig:abchpdm2}
\end{figure}

>>>>

\begin{figure}
    \includegraphics[width=\textwidth]{abc-boxplot-mN}
    \caption[
        Posterior mean point estimates for BA model parameters $m$ and $N$
        obtained by running \textit{netabc} on simulated data, stratified by
        true parameter values.
    ]{
        Posterior mean point estimates for BA model parameters $m$ and $N$
        obtained by running \textit{netabc} on simulated data, stratified by
        true parameter values. First row of plots contains true versus
        estimated values of $m$; second row contains true versus estimated
        values of $N$. Columns are stratified by $\alpha$, $I$, and $m$
        respectively. Dashed lines indicate true values. 
    }
    \label{fig:abcpt2}
\end{figure}

%\begin{figure}[ht]
%    \centering
%    \includegraphics{abc-point-estimate-m3.pdf}
%    \caption[
%        Posterior mean point estimates for \gls{BA} model parameters obtained
%        by running \software{netabc} on simulated data, for simulations with
%        \gls{m} = 3. 
%    ]{
%        Posterior mean point estimates for \gls{BA} model parameters obtained
%        by running \software{netabc} on simulated data, for simulations with
%        \gls{m} = 3. Dashed lines indicate true values. (A) Estimates of
%        \gls{alpha} and \gls{I} which were varied in these simulations against
%        known values. (B) Estimates of \gls{m} and \gls{N} which were held
%        fixed in these simulations at the values \gls{m} = 3 and \gls{N} =
%        5000.
%    }        
%    \label{fig:abcptm3}
%\end{figure}
%
%\begin{figure}[ht]
%    \centering
%    \includegraphics{abc-point-estimate-m4.pdf}
%    \caption[
%        Posterior mean point estimates for \gls{BA} model parameters obtained
%        by running \software{netabc} on simulated data, for simulations with
%        \gls{m} = 3.
%    ]{
%        Posterior mean point estimates for \gls{BA} model parameters obtained
%        by running \software{netabc} on simulated data, for simulations with
%        \gls{m} = 4. Dashed lines indicate true values. (A) Estimates of
%        \gls{alpha} and \gls{I} which were varied in these simulations against
%        known values. (B) Estimates of \gls{m} and \gls{N} which were held
%        fixed in these simulations at the values \gls{m} = 4 and \gls{N} =
%        5000.
%    }        
%    \label{fig:abcptm4}
%\end{figure}

@@

%\begin{figure}[ht]
%    \centering
%    \includegraphics{mixed-posterior}
%    \caption[
%        Approximate marginal posterior distributions of \gls{BA} model
%        parameters obtained with \software{netabc} for a network with
%        heterogeneous node behaviour.
%    ]{
%        Approximate marginal posterior distributions of \gls{BA} model
%        parameters obtained with \software{netabc} for a network with
%        heterogeneous node behaviour. Half of the nodes were attached with
%        $\gls{alpha} = 0.5$, and the other half with $\gls{alpha} = 1.5$
%        (vertical dashed lines, top left). Other parameter values were $m = 2$,
%        $\gls{I} = 1000$, and $\gls{N} = 5000$ (vertical dashed lines, other
%        than top left). Shaded areas indicate 95\% highest posterior density
%        intervals.
%    }
%    \label{fig:mixed}
%\end{figure}
%
%%\begin{figure}[ht]
%%    \centering
%%    \includegraphics{peerdriven-posterior}
%%    \caption[
%%        Approximate marginal posterior distributions of \gls{BA}
%%        model parameters obtained with \software{netabc} for a network with
%%        peer-driven sampling.
%%    ]{
%%        Approximate marginal posterior distributions of \gls{BA}
%%        model parameters obtained with \software{netabc} for a network with
%%        peer-driven sampling. An epidemic was simulated in the usual fashion,
%%        but rather than being sampled at random, infected nodes were sampled
%%        with a probability two times higher if they had any sampled neighbours
%%        in the contact network. Vertical dashed lines indicate true parameter
%%        values, and shaded areas indicate 95\% highest posterior density
%%        intervals.
%%    }
%%    \label{fig:peerdriven}
%%\end{figure}
%
\begin{figure}[ht]
    \includegraphics{realdata-hpd-bc-m2}
    \caption[
        Posterior mean point estimates and 95\% \gls{HPD} intervals for
        parameters of the \gls{BA} network model, fitted to eleven \gls{HIV}
        datasets with \software{netabc} using the prior \gls{m} $\sim$
        DiscreteUniform(2, 5).
    ]{
        Posterior mean point estimates and 95\% \gls{HPD} intervals for
        parameters of the \gls{BA} network model, fitted to five published
        \gls{HIV} datasets with \software{netabc} using the prior $\gls{m}
        \sim \DiscreteUniform(2, 5)$. $x$-axes indicate regions of nonzero
        prior density. 
    }
    \label{fig:abchpdm2}
\end{figure}

### 2016-07-27 doc/thesis/discussion.Rnw
Defence version of thesis with changes

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered
(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
\gls{alpha} values produce networks with few well-connected ``superspreader''
nodes which are involved in a large number of transmissions, resulting in a
highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
\del{There was no observable bias in the estimates of \gls{alpha} obtained with
\software{netabc}, however the variation in these estimates
was higher for $\alpha < 1$ than for $\alpha \geq 1$} \add{There appeared to be
weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
(\cref{fig:abcptm2,tab:abchpd}), which may be partially explained by the
relationship between \gls{alpha} and the power law exponent \gls{gamma}
(\cref{fig:gamma}). \add{Although the degree distributions do not truly follow
a power law for $\gls{alpha} \neq 1$, the fitted exponent still captures the
slope of the degree distribution reasonably well (\cref{fig:powerlaw}), and 
distributions with similar fitted \gls{gamma} values appear qualitatively
similar in other respects.} The \gls{gamma} values fitted to $\gls{alpha} = 0$
and $\gls{alpha} = 0.5$ are nearly identical (about $2.28 for \gls{alpha} = 0$
and $2.33 for \gls{alpha} = 0.5$ with $\gls{N} = 5000$ and $\gls{m} = 2$). In
other words, the degree distributions of networks with \gls{alpha} $< 1$ are
similar to each other, which may result in similarity of corresponding
transmission trees as well.

@@

The number of nodes in the network, \gls{N}, exhibited the most variation in
terms of its effect on tree shape. There was almost no difference between trees
simulated under different \gls{N} values when the number of infected nodes
\gls{I} was small (\cref{fig:Nkpca,fig:Ncrossv}). There is an intuitive
explanation for this result, namely that adding additional nodes does not
change the edge density or overall shape of a \gls{BA} network. \add{In fact,
this is why the degree distributions of these networks are called
``scale-free.''} This can be illustrated by imagining that we add a small
number of nodes to a network after the epidemic simulation has already been
completed. It is possible that none of these new nodes attains a connection to
any infected node. Thus, running the simulation again on the new, larger
network could produce the exact same transmission tree as before. On the other
hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
discussed above also apply. The waiting times until the next infection
increase, resulting in longer coalescence times toward the tips. The relative
accuracy of the \gls{nltt} in these situations
(\cref{fig:rsquared,fig:Ncrossv}) corroborates this hypothesis, as the
\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
parameters were simultaneously estimated with \gls{ABC}, \gls{N} was nearly
always over-estimated by approximately a factor of two
(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
bias was our choice of prior distribution. Since the prior for \gls{I} and
\gls{N} was jointly uniform on a region where $I \leq N$, more prior weight was
assigned to higher \gls{N} values. Another contributing factor relates to the
dynamics of the \gls{SI} model and the coalescent process and is discussed
below.

@@

The combination of method, model, and priors we employed did not produce
perfect estimates of any of the parameters. The estimates of \gls{alpha} were
the most accurate, although the variance of the estimates was high and the
confidence intervals were wide for $\gls{alpha} < 1$
(\cref{tab:abchpd,fig:abcptm2,fig:abcptm3,fig:abcptm4}). The estimates of
\gls{N} and \gls{I} were both upwardly biased, and the estimates of \gls{m}
were largely uninformative. Despite these limitations, a major result of our
investigation is that some contact network parameters have a measurable impact
on tree shape which can be used to perform statistical inference. Further
refinements to \software{netabc}, as well as the use of more sophisticated
network models, may improve the accuracy and precision of these estimates.

@@

\add{Unfortunately, the sub-linear region for \gls{alpha} identified by both
\textcite{de2007preferential} and \software{netabc} is also the region of
poorest identifiability (\cref{fig:gridest,fig:abcptm2}). This was reflected in
the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
intervals covering the majority of the range [0, 1]. The value $\gls{alpha} =
0.5$ was contained in the 95\% \gls{HPD} interval for every dataset;
consequently, it is not possible to say with high confidence that any of the
\gls{alpha} values are different from each other. In synthetic data, the
confidence intervals around \gls{alpha} narrowed when other parameters were
marginalized out (\cref{fig:marginal}). Thus, it is possible that estimates of
\gls{alpha} could be made more precise by specifying either exact values or
informative priors on the other \gls{BA} parameters. We argued informally in
\cref{subsec:whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
estimated by standard epidemiological methods, although the difficulty may
depend on the specific epidemic under consideration.}

@@

\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
populations in sub-Saharan Africa, where the dominant mode of transmission is
believed to be heterosexual. The HET/Uganda study population included
1434 seropositive individuals. Since the majority of inhabitants of the study
district were enrolled in the study, it is likely that this number represents
the majority of \gls{HIV}-infected persons. Our estimate was slightly lower
    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(mean)]}),
but the value 1434 was contained within the 95\% \gls{HPD} interval
    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.max.95)]}).
The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
sequencing effort, carried out on samples collected for other purposes between
1981 and 1989. Every individual in the study district was sampled; among these,
a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
lowest estimated \gls{I} value among the heterosexual datasets, 
the posterior mean
    (\Sexpr{d1[dataset == "mccormack2002" & parameter == "I", round(mean)]}),
was much higher than this value. Due to the thorough population-level
screening, it is most likely that the posterior mean represents an
overestimate, especially considering the observed upward bias of \gls{ABC}
estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
It is also possible that the true sexual network extended outside the study's
geographic area to include a larger number of infected individuals. The authors
of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
individuals in the study area. \gls{HIV} sequences were obtained from
approximately 70\% of these individuals. The estimated prevalence we obtained
was much higher 
    (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
authors noted that the town where the study took place was located proximally
to the country's capital city, and suggested that frequent travel between the
two locations may have facilitated linking of their sexual networks. Thus, the
high estimate of \gls{I} we obtained for these data may include a larger
network component based in the capital city.}

>>>>

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered
(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
\gls{alpha} values produce networks with few well-connected ``superspreader''
nodes which are involved in a large number of transmissions, resulting in a
highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
\del{There was no observable bias in the estimates of \gls{alpha} obtained with
\software{netabc}, however the variation in these estimates
was higher for $\alpha < 1$ than for $\alpha \geq 1$} \add{There appeared to be
weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
(\cref{fig:abcpt,tab:abchpd}), which may be partially explained by the
relationship between \gls{alpha} and the power law exponent \gls{gamma}
(\cref{fig:gamma}). \add{Although the degree distributions do not truly follow
a power law for $\gls{alpha} \neq 1$, the fitted exponent still captures the
slope of the degree distribution reasonably well (\cref{fig:powerlaw}), and 
distributions with similar fitted \gls{gamma} values appear qualitatively
similar in other respects.} The \gls{gamma} values fitted to $\gls{alpha} = 0$
and $\gls{alpha} = 0.5$ are nearly identical (about $2.28 for \gls{alpha} = 0$
and $2.33 for \gls{alpha} = 0.5$ with $\gls{N} = 5000$ and $\gls{m} = 2$). In
other words, the degree distributions of networks with \gls{alpha} $< 1$ are
similar to each other, which may result in similarity of corresponding
transmission trees as well.

@@

The number of nodes in the network, \gls{N}, exhibited the most variation in
terms of its effect on tree shape. There was almost no difference between trees
simulated under different \gls{N} values when the number of infected nodes
\gls{I} was small (\cref{fig:Nkpca,fig:Ncrossv}). There is an intuitive
explanation for this result, namely that adding additional nodes does not
change the edge density or overall shape of a \gls{BA} network. \add{In fact,
this is why the degree distributions of these networks are called
``scale-free.''} This can be illustrated by imagining that we add a small
number of nodes to a network after the epidemic simulation has already been
completed. It is possible that none of these new nodes attains a connection to
any infected node. Thus, running the simulation again on the new, larger
network could produce the exact same transmission tree as before. On the other
hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
discussed above also apply. The waiting times until the next infection
increase, resulting in longer coalescence times toward the tips. The relative
accuracy of the \gls{nltt} in these situations
(\cref{fig:rsquared,fig:Ncrossv}) corroborates this hypothesis, as the
\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
parameters were simultaneously estimated with \gls{ABC}, \gls{N} was nearly
always over-estimated by approximately a factor of two
(\cref{fig:abcpt2,tab:abchpd}). One factor which may have contributed to this
bias was our choice of prior distribution. Since the prior for \gls{I} and
\gls{N} was jointly uniform on a region where $I \leq N$, more prior weight was
assigned to higher \gls{N} values. Another contributing factor relates to the
dynamics of the \gls{SI} model and the coalescent process and is discussed
below.

@@

The combination of method, model, and priors we employed did not produce
perfect estimates of any of the parameters. The estimates of \gls{alpha} were
the most accurate, although the variance of the estimates was high and the
confidence intervals were wide for $\gls{alpha} < 1$
(\cref{fig:abcpt,tab:abchpd}). The estimates of \gls{N} and \gls{I} were both
upwardly biased, and the estimates of \gls{m} were largely uninformative.
Despite these limitations, a major result of our investigation is that some
contact network parameters have a measurable impact on tree shape which can be
used to perform statistical inference. Further refinements to
\software{netabc}, as well as the use of more sophisticated network models, may
improve the accuracy and precision of these estimates.

@@

\add{Unfortunately, the sub-linear region for \gls{alpha} identified by both
\textcite{de2007preferential} and \software{netabc} is also the region of
poorest identifiability (\cref{fig:gridest,fig:abcpt}). This was reflected in
the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
intervals covering the majority of the range [0, 1]. The value $\gls{alpha} =
0.5$ was contained in the 95\% \gls{HPD} interval for every dataset;
consequently, it is not possible to say with high confidence that any of the
\gls{alpha} values are different from each other. In synthetic data, the
confidence intervals around \gls{alpha} narrowed when other parameters were
marginalized out (\cref{fig:marginal}). Thus, it is possible that estimates of
\gls{alpha} could be made more precise by specifying either exact values or
informative priors on the other \gls{BA} parameters. We argued informally in
\cref{subsec:whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
estimated by standard epidemiological methods, although the difficulty may
depend on the specific epidemic under consideration.}

@@

\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
populations in sub-Saharan Africa, where the dominant mode of transmission is
believed to be heterosexual. The HET/Uganda study population included
1434 seropositive individuals. Since the majority of inhabitants of the study
district were enrolled in the study, it is likely that this number represents
the majority of \gls{HIV}-infected persons. Our estimate was slightly lower
    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(mean)]}),
but the value 1434 was contained within the 95\% \gls{HPD} interval
    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.min.95)]} -
    \Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.max.95)]}).
The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
sequencing effort, carried out on samples collected for other purposes between
1981 and 1989. Every individual in the study district was sampled; among these,
a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
lowest estimated \gls{I} value among the heterosexual datasets, 
the posterior mean
    (\Sexpr{d1[dataset == "mccormack2002" & parameter == "I", round(mean)]}),
was much higher than this value. Due to the thorough population-level
screening, it is most likely that the posterior mean represents an
overestimate, especially considering the upward bias of \gls{ABC} estimates of
\gls{I} observed on simulated data (\cref{fig:abcpt2}). It is also possible
that the true sexual network extended outside the study's geographic area to
include a larger number of infected individuals. The authors of the
HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
estimated that there were 1731 \gls{HIV}-positive individuals in the study
area. \gls{HIV} sequences were obtained from approximately 70\% of these
individuals. The estimated prevalence we obtained was much higher 
    (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
authors noted that the town where the study took place was located proximally
to the country's capital city, and suggested that frequent travel between the
two locations may have facilitated linking of their sexual networks. Thus, the
high estimate of \gls{I} we obtained for these data may include a larger
network component based in the capital city.}
